Basics of Hardware Cache Optimization

### 2017-06-08 tex/list-vs-vector.tex
Expand, revise, tweak

\alts{%
   The C++ program shown in \cref{lst:ithare},
   The following C++ program,
}
\x{, adapted from \textcite{bigos},} initializes a number of \acrshort{stl} containers
with random numbers and measures the processor time needed to \alts{sum all of them,
compute the sum}.
%
I first ran it with \mintinline{cpp}{Container} being a type alias for
\mintinline{cpp}{std::list}, then for \mintinline{cpp}{std::vector}.%
\footnote{%
   Each compiled with \x{the \gls{gcc}} and \mintinline{text}{-O3} and
   \mintinline{text}{-march=native}
}

\begin{center}
   \inputminted[firstline=20]{cpp}{ithare/list-vs-vector.cpp}
   \captionof{listing}{Compute the Sum of a Container, adapted from \textcite{bigos}}
   \label{lst:ithare}
\end{center}

@@

When using \mintinline{cpp}{std::vector}, computing the sum is
\num[round-mode=places, round-precision=0]{\speedup} times faster.

>>>>

\alts{%
   The C++ program shown in \cref{lst:ithare},
   The following C++ program,
}%
, adapted from \textcite{bigos}, initializes a number of \acrshort{stl} containers
with random numbers and measures the processor time needed to \alts{sum all of them,
compute the sum}.
%
I first ran it with \mintinline{cpp}{Container} being a type alias for
\mintinline{cpp}{std::list}, then for \mintinline{cpp}{std::vector}.%
\footnote{%
   Each compiled with \x{the \gls{gcc}} and \mintinline{text}{-O3} and
   \mintinline{text}{-march=native}
}
Either way, the asymptotic complexity is \(\Theta(N)\).

\begin{center}
   \inputminted[firstline=29]{cpp}{ithare/list-vs-vector.cpp}
   \captionof{listing}{Compute the Sum of a Container, adapted from \textcite{bigos}}
   \label{lst:ithare}
\end{center}

@@

% > In trying to explain this difference, we'll need to get into educated-guesswork area.
% > For MSVC and gcc, the performance difference between vector<> and list<> is pretty
% > much in line with the difference between typical cached access times (single-digit
% > clocks) and typical uncached access times (100-150 clocks).  As access patterns for
% > vector<> are expected to use CPU prefetch fully and list<> under the patterns in
% > Listing 1 is pretty much about random access to memory, which cannot be cached due to
% > the size, this 100-150× difference in access times can be expected to translate into
% > 100-150× difference in performance.
%      -- https://accu.org/index.php/journals/2268

% > [T]he performance difference between vector<> and list<> is pretty
% > much in line with the difference between typical cached [...] and typical uncached
% > access times.
%      -- https://accu.org/index.php/journals/2268

% \noindent
% \alts{{The result is that, when}, When}  computing the
% sum \alts{runs, is} \num[round-mode=places, round-precision=0]{\speedup} times faster.
\alts{My, The} result is that computing the sum \alts{runs, is} \num[round-mode=places,
round-precision=0]{\speedup} times faster when using \mintinline{cpp}{std::vector}.  Some
of this difference can certainly be attributed to space overhead of the linked list and
the added indirection, but the more \emph{cache-friendly} memory access pattern of
\mintinline{cpp}{std::vector} is \alts{decisive, key, crucial, vital}~\cite[5\psq]{bigos}.
Using \mintinline{cpp}{std::list} incurs \say{random access to memory} in this
example~\cite[6]{bigos}.

### 2017-06-08 tex/methodology.tex
Expand, revise, tweak

All programs were compiled and run on Linux using the \gls{gcc} with
\mintinline{text}{-O2} and \mintinline{text}{-march=native}.
% Additionally,
I disabled CPU frequency scaling with \mintinline{text}{cpupower(1)},
% See <https://wiki.archlinux.org/index.php/CPU_frequency_scaling>.
reduced the number of running processes,%
\footnote{%
   I stopped Firefox (which added visible noise) and the X server, for example.
}
% (e.g. leaving Firefox open in the background added lots of noise),
and assigned a high priority to the benchmark process with the \mintinline{text}{chrt(1)}
command.

The \emph{\href{https://github.com/meribold/cache-seminar-paper/blob/master/makefile}
{makefile}} used to compile and run the measurements as well as the full source code of
the programs is available \alts{here, on GitHub}:
\url{https://github.com/meribold/cache-seminar-paper}.

>>>>

All programs were compiled and run on Linux using the \gls{gcc} with
\mintinline{text}{-O2} (unless noted otherwise) and \mintinline{text}{-march=native}.
% Additionally,
I disabled CPU frequency scaling with \mintinline{text}{cpupower(1)},
% See <https://wiki.archlinux.org/index.php/CPU_frequency_scaling>.
reduced the number of running processes,%
\footnote{%
   I stopped Firefox (which added visible noise) and the X server, for example.
}
% (e.g. leaving Firefox open in the background added lots of noise),
and assigned a high priority to the benchmark process with the \mintinline{text}{chrt(1)}
command.

The \emph{\href{https://github.com/meribold/cache-seminar-paper/blob/master/makefile}
{makefile}} used to compile and run the measurements as well as the full source code of
the programs is available \alts{at, here:, on GitHub:}
\url{https://github.com/meribold/cache-seminar-paper}.
